---
title: "Portfolio"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(ggplot2)
library(flexdashboard)
library(compmus)
library(spotifyr)
```

```{r include=FALSE}
Clipping <- get_artist_audio_features('5HJ2kX5UTwN4Ns8fB5Rn1I')
Jpegmafia <- get_artist_audio_features('6yJ6QQ3Y5l0s0tn7b0arrO')
Death_Grips <- get_artist_audio_features('5RADpgYLOuS2ZxDq7ggYYH')
```

### Introduction

This portfolio aims to analyze a specific branch of hip-hop, namely industrial hip-hop. Industrial hip-hop, in my opinion, is one of the most interesting and bizarre genres in rap music. Being a fusion of industrial music and hip-hop, the genre showcases artists blending harsh, mechanical, transgressive or provocative sounds accompanied with rap and rhythm. Due to being a niche genre, one which many people may not be familiar with, I believe it would be a very interesting topic to explore in this online portfolio. The aim of the portfolio is to research how industrial hip-hop formed as a unique branch in hip-hop and, more specifically, to analyze songs in the genre from the 80’s to the current day to see what elements they have in common, and in which way these songs differ and have evolved. Elements such as loudness, lyrics, timbre, and bpm would be interesting to explore over time as I believe there are visible changes in these areas throughout the years, such as loudness and bpm increasing. The portfolio would possibly also include comparing industrial hip-hop to other branches of rap music in order to better understand what elements in a song make it uniquely industrial and set it apart from its counterparts.

Artists I wish to analyze for this portfolio include the following:

2000’s-present: Jpegmafia, Clipping, Death Grips. These are all artists I personally listen to and all bring their own unique sound to the industrial hip-hop genre.

1990’s: Young Black Tennagers, Scorn. These I believe can show more about the forming of more experimentation with industrial noise in rap music.

1980’s: Mark Stewart, Bill Laswell, and Adrian Sherwood. These artists appear to have set the roots for industrial hip-hop, and so analyzing these songs I believe will reveal more about the forming of the genre itself.

### Current Plots

```{r echo=FALSE}
data = rbind(Clipping, Jpegmafia, Death_Grips)
ggplot(data, aes(x=energy, y=loudness, color = artist_name)) + 
geom_point() + ggtitle("Current industrial hip-hop artists' songs measured by Loudness and Energy")
```

```{r echo=FALSE}
ggplot(data, aes(x=danceability, y=tempo, color = artist_name)) + 
geom_point() + ggtitle("Current industrial hip-hop artists' songs measured by tempo and danceability")
```

------------------------------------------------------------------------

The current plots show the relations between loudness/energy and tempo/dance. The current goal of these plots is to find what common elements these artists within the genre tend to share and which elements vastly differ. The hope is then to get a better understanding of what makes a hip-hop song industrial. From there we can start comparing songs of the genre with other genres which may or may not have inspired it and see the evolution of song elements.

### Conclusion/discussion
